\NAME: A framework for \\ Efficient Intrusion Detection\\ in the Internet of Things

### 2014-11-24 paper.tex
fixed various typos, bad sentences,…

When intrusion prevention fails, intrusion detection serves as a second layer
of defense. It looks for patterns \& anomalies that can indicate malicious
behavior. However, supporting an adequate set of intrusion detection algorithms
imposes significant overhead on resource-constrained devices such as those in
the Internet of Things. Fusion of the algorithms reduces resource consumption
by optimizing resource sharing, yet, proves to be time-consuming, repetitive
and error-prone. To address this problem we propose \NAME, a framework for the
development of efficient intrusion detection systems. It consists of a domain
specific language, to formally describe intrusion detection algorithms on a
functional level, and a code generator to produce more optimally organized
source code. A side-by-side comparison shows that \NAME-generated code is
smaller and reduces message passing overhead, execution time and memory
footprint in comparison to a straightforward sequential implementation of the
same algorithms.

@@

Current developments in ID focus on programming frameworks that structure the
implementation of algorithms \cite{valero2012di} and offer the required basic
functional components to implement them \cite{krontiris2008lidea}. A key
shortcoming of prior ID frameworks, is that they don't offer a way to optimize
the usage of resources. A solution that supports the integration of multiple
algorithms on IoT devices needs to avoid accumulating their impact on the
available resources. A second omission is support for systematic reuse to
create different configurations heterogeneous devices. Addressing this in a
transparent and automated way offers the potential to significantly reduce the
effort of developing an IDS.

@@

The first contribution of this paper is a framework called \NAME. It enables
the formal description of ID algorithms on a functional level and provides a
code generator to produces corresponding source code for a given platform and
configuration. This way, \NAME addresses the heterogeneous nature of both the
IoT and ID algorithms.

@@

The second contribution introduces functional code fusion (FCF) as a source
code generation paradigm to solve the problem of hard to reuse and combine ID
algorithms. FCF identifies common data and functions, and organizes code to
eliminate redundant iterations, tests and computations. Results show improved
image size, reduced usage of the wireless radio, reduced execution time and
lower memory usage.

@@

Using a platform independent language also offers the possibility to relief the
implementer of explicit memory management and low-level data structures. From a
language perspective, \NAME offers several programming constructs, typically
found in modern programming languages with higher levels of abstraction:
optional typing, object orientation, pattern matching, aspect orientation.

@@

To convert the DSL-based ID algorithms into production-ready code, \NAME
applies code generation techniques. Although different approaches here are
applicable, like template-based text expansion, XSLT transformations or CodeDom
\cite{dollard2004code}, we believe that the functional transformations require
a richer model.

@@

In this section we present a prototype implementation and evaluates its
performance. We first present the implementation consisting of the hardware
platform, network and software stack. Next we use this implementation to
quantify its merits using a side-by-side comparison of generated code versus a
manual implementation. We consider experimental results regarding the image
size, usage of the network and execution time of the event-loop. We also take a
more analytical approach and look at the handling of common data and the size
of the source code.

@@

We evaluated three criteria: the network usage in number of packets and bytes,
the time required to perform once cycle of the event loop and the image size of
the resulting compiled code. To do this, we collected these metrics for four
situations: without ID algorithms, with a watchdog, with reputation-tracking
and with both algorithms implemented.

@@

Looking at the generated source code is another way to validate the framework.
A simple metric is comparing the amount of code versus the size of the formal
description of IDS algorithms. This can be an indication of the development
effort and complexity. In this experiment, implementing both algorithms
manually results in 637 lines of C code. The corresponding \NAME implementation
results in 156 lines of DSL code. This is largely thanks to the higher level
coding constructs, as introduced in section \ref{language-support} dealing with
\NAME's language support.

@@

This section focuses on related work in the field of domain specific languages
and code generation, both in relation to wireless networks of
resource-constraint devices. These two components are the core of the \NAME
framework and the solution it proposes. We also look at their application in
other disciplines and how this can relate to our context.

@@

Within the scope of wireless networks of resource-constraint devices, we see
the introduction of DSLs \enquote{to shorten the development cycles
dramatically during deployments, to reduce the dependency on hardware and
software engineers, and to lead to a wider adoption outside the computer
science field} \cite{sadilek2008domain}.

@@

DSLs offer a solution here, but other paradigms are possible. A mixed framework
consisting of a virtual machine (VM) and native code extension
\cite{sadilek2007energy}, combined with energy-aware compilation, seems a
promising approach. By offloading the optimization to a VM and only compiling
parts and not the whole, the solution allows updating at a granular scale, thus
reducing the update cost, while maintaining flexibility and lowering the
development effort. As explained in section \ref{positioning}, \NAME focuses on
the reduction of the development effort and update cost, while preserving
maximal flexibility.

@@

Adding multiple intrusion detection algorithms to wireless resource-constraint
devices requires fusion of these algorithms on a functional level. This can
reduce execution inefficiencies, redundant memory usage and superfluous network
access.

>>>>

When intrusion prevention fails, intrusion detection serves as a second layer
of defense. It looks for patterns and anomalies that can indicate malicious
behavior. However, supporting an adequate set of intrusion detection algorithms
imposes significant overhead on resource-constrained devices such as those in
the Internet of Things. Fusion of the algorithms reduces resource consumption
by optimizing resource sharing, yet, proves to be time-consuming, repetitive
and error-prone. To address this problem we propose \NAME, a framework for the
development of efficient intrusion detection systems. It consists of a domain
specific language, to formally describe intrusion detection algorithms on a
functional level, and a code generator to produce more optimally organized
source code. A side-by-side comparison shows that \NAME-generated code is
smaller and reduces message passing overhead, execution time and memory
footprint in comparison to a straightforward sequential implementation of the
same algorithms.

@@

Current developments in ID on resource-constrained devices focus on
programming frameworks that structure the implementation of algorithms
\cite{valero2012di} and offer the required basic functional components to
implement them \cite{krontiris2008lidea}. A key shortcoming of prior ID
frameworks, is that they don't offer a way to optimize the usage of resources.
A solution that supports the integration of multiple algorithms on IoT devices
needs to avoid accumulating their impact on the available resources. A second
omission is support for systematic reuse to create different configurations
heterogeneous devices. Addressing this in a transparent and automated way
offers the potential to significantly reduce the effort of developing an IDS.

@@

The first contribution of this paper is a framework called \NAME. It enables
the formal description of ID algorithms on a functional level and provides a
code generator to produce corresponding source code for a given platform and
configuration. This way, \NAME addresses the heterogeneous nature of both the
IoT and ID algorithms.

@@

The second contribution introduces functional code fusion (FCF) as a source
code generation paradigm to solve the problem of hard to reuse and combine ID
algorithms. FCF identifies common data and functions, and organizes code to
eliminate redundant iterations, tests and computations. Results show reduced
usage of the wireless radio, reduced execution time, lower memory usage and
improved image size.

@@

Using a platform independent language also offers the possibility to relieve
the implementer of explicit memory management and low-level data structures.
From a language perspective, \NAME offers several programming constructs,
typically found in modern programming languages with higher levels of
abstraction: optional typing, object orientation, pattern matching, aspect
orientation.

@@

To convert the DSL-based ID algorithms into production-ready code, \NAME
applies code generation techniques. Although different approaches here are
applicable, like template-based text expansion, XSLT transformations or CodeDom
\cite{dollard2004code}, we believe that the required transformations demand a
richer model.

@@

In this section we present a prototype implementation and evaluate its
performance. We first present the implementation consisting of the hardware
platform, network and software stack. Next we use this implementation to
quantify its merits using a side-by-side comparison of generated code versus a
manual implementation. We consider experimental results regarding usage of the
network, execution time of the event-loop and the image size. We also take a
more analytical approach and look at the handling of common data and the size
of the source code.

@@

We evaluated four criteria: the network usage in number of packets and bytes,
the time required to perform once cycle of the event loop and the image size of
the resulting compiled code. To do this, we collected these metrics for four
situations: without ID algorithms, with a watchdog, with reputation-tracking
and with both algorithms implemented.

@@

Looking at the generated source code is another way to validate the framework.
A simple metric is comparing the amount of code versus the size of the formal
description of ID algorithms. This can be an indication of the development
effort and complexity. In this experiment, implementing both algorithms
manually results in 637 lines of C code. The corresponding \NAME implementation
results in 156 lines of DSL code. This is largely thanks to the higher level
coding constructs, as introduced in section \ref{language-support} dealing with
\NAME's language support.

@@

This section focuses on related work in the field of domain specific languages
and code generation, both in relation to wireless networks of
resource-constrained devices. These two components are the core of the \NAME
framework and the solution it proposes. We also look at their application in
other disciplines and how this can relate to our context.

@@

Within the scope of wireless networks of resource-constrained devices, we see
the introduction of DSLs \enquote{to shorten the development cycles
dramatically during deployments, to reduce the dependency on hardware and
software engineers, and to lead to a wider adoption outside the computer
science field} \cite{sadilek2008domain}.

@@

DSLs offer a solution here, but other paradigms are possible. A mixed framework
consisting of a virtual machine (VM) and native code extension
\cite{sadilek2007energy}, combined with energy-aware compilation, seems a
promising approach. By offloading the optimization to a VM and only compiling
parts and not the whole, the solution allows updating at a granular scale, thus
reducing the update cost, while maintaining flexibility and lowering the
development effort. In this terminology, \NAME focuses on the reduction of the
development effort and update cost, and maximizes flexibility.

@@

Adding multiple intrusion detection algorithms to wireless resource-constrained
devices requires fusion of algorithms on a functional level. This can reduce
execution inefficiencies, redundant memory usage and superfluous network access.
